---
title: "01_data-cleaning"
format: pdf
---

```{r}
# loading libraries

library(tidyverse)
```


```{r}
# loading datasets

interarrival <- read_csv("data/01_interarrival_by_timeblock_hourly_cy2025.csv")
case_mix <- read_csv("data/02_case_mix_by_timeblock_cy2025.csv")
first_seen_summary <- read_csv("data/03_arrival_to_first_seen_distribution_by_triage_cy2025.csv")
first_seen_empirical <- read_csv("data/04_arrival_to_first_seen_empirical_deid_cy2025.csv")
workup_summary <- read_csv("data/05_generic_workup_duration_distribution_by_complexity_cy2025.csv")
workup_empirical <- read_csv("data/06_generic_workup_duration_empirical_deid_cy2025.csv")
imaging_prob <- read_csv("data/07_imaging_probability_and_modality_mix_by_complexity_acuity_cy2025.csv")
imaging_duration <- read_csv("data/08_imaging_duration_distribution_by_modality_cy2025.csv")
```

### EDA
```{r}
glimpse(interarrival)
glimpse(case_mix)
glimpse(first_seen_summary)
glimpse(first_seen_empirical)
glimpse(workup_summary)
glimpse(workup_empirical)
glimpse(imaging_prob)
glimpse(imaging_duration)
```

```{r}
head(interarrival)
head(case_mix)
head(first_seen_summary)
head(first_seen_empirical)
head(workup_summary)
head(workup_empirical)
head(imaging_prob)
head(imaging_duration)
```
```{r}
check_missing <- function(df, name) {
  cat("\n========================\n")
  cat("Missing values in:", name, "\n")
  print(colSums(is.na(df)))
}

check_missing(interarrival, "interarrival")
check_missing(case_mix, "case_mix")
check_missing(first_seen_summary, "first_seen_summary")
check_missing(first_seen_empirical, "first_seen_empirical")
check_missing(workup_summary, "workup_summary")
check_missing(workup_empirical, "workup_empirical")
check_missing(imaging_prob, "imaging_prob")
check_missing(imaging_duration, "imaging_duration")
```
1. Interarrival: This file controls when patients arrive into the simulation. The README says to use it to drive entity generation.
```{r}
# Interarrival EDA

# How many rows / unique hours / day types?
nrow(interarrival)
unique(interarrival$day_of_week_label)
sort(unique(interarrival$hour_of_day))

# Average arrival rate by hour of day
arrival_by_hour <- interarrival %>%
  group_by(hour_of_day) %>%
  summarise(
    mean_arrival_rate = mean(arrival_rate_per_hour, na.rm = TRUE),
    mean_interarrival = mean(mean_interarrival_min, na.rm = TRUE)
  )

print(arrival_by_hour)

# Plot: arrival rate by hour
ggplot(arrival_by_hour, aes(x = hour_of_day, y = mean_arrival_rate)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Arrival Rate by Hour of Day",
    x = "Hour of Day",
    y = "Average Arrival Rate per Hour"
  )

# Plot: mean interarrival time by hour
ggplot(arrival_by_hour, aes(x = hour_of_day, y = mean_interarrival)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Mean Interarrival Time by Hour of Day",
    x = "Hour of Day",
    y = "Mean Interarrival Time (minutes)"
  )

# By day of week and hour
arrival_heatmap <- interarrival %>%
  group_by(day_of_week_label, hour_of_day) %>%
  summarise(mean_rate = mean(arrival_rate_per_hour, na.rm = TRUE), .groups = "drop")

ggplot(arrival_heatmap, aes(x = hour_of_day, y = day_of_week_label, fill = mean_rate)) +
  geom_tile() +
  labs(
    title = "Arrival Rate Heatmap by Day of Week and Hour",
    x = "Hour of Day",
    y = "Day of Week",
    fill = "Arrival Rate"
  )

```
2. Case Mix: This file tells you what kind of patient arrives.
```{r}
# Case mix EDA

# How many values for each attribute?
case_mix %>%
  group_by(attribute_name) %>%
  summarise(n_values = n_distinct(attribute_value))

# Example: look at probabilities for each attribute
case_mix %>%
  group_by(attribute_name, attribute_value) %>%
  summarise(avg_prob = mean(probability, na.rm = TRUE), .groups = "drop") %>%
  arrange(attribute_name, desc(avg_prob))



```

3. Arrival to First Seen (summary): This file represents the front-end delay from arrival until first provider contact.
```{r}
# Arrival to first seen

# Look at delays by triage priority
print(first_seen_summary)

# Bar plot of median first-seen time by triage priority
ggplot(first_seen_summary, aes(x = factor(triage_priority), y = median_min)) +
  geom_col() +
  labs(
    title = "Median Arrival-to-First-Seen Time by Triage Priority",
    x = "Triage Priority",
    y = "Median Minutes"
  )

# p90 comparison
ggplot(first_seen_summary, aes(x = factor(triage_priority), y = p90_min)) +
  geom_col() +
  labs(
    title = "90th Percentile Arrival-to-First-Seen Time by Triage Priority",
    x = "Triage Priority",
    y = "P90 Minutes"
  )
```
4. Arrival to First Seen (empirical): This file has de-identified encounter-level durations rather than just summary statistics.
```{r}
# Empirical first seen


# Histogram
ggplot(first_seen_empirical, aes(x = duration_min)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of Arrival-to-First-Seen Durations",
    x = "Minutes",
    y = "Count"
  )

# By triage priority
ggplot(first_seen_empirical, aes(x = factor(triage_priority), y = duration_min)) +
  geom_boxplot() +
  labs(
    title = "Arrival-to-First-Seen Duration by Triage Priority",
    x = "Triage Priority",
    y = "Minutes"
  )
```

5. Generic Workup Summary: This is the big middle chunk of the ED visit
```{r}
# Workup summary


# Median workup time by complexity
ggplot(workup_summary, aes(x = complexity_bucket, y = median_min)) +
  geom_col() +
  labs(
    title = "Median Generic Workup Duration by Complexity",
    x = "Complexity Bucket",
    y = "Median Minutes"
  )

# p90 workup time by complexity
ggplot(workup_summary, aes(x = complexity_bucket, y = p90_min)) +
  geom_col() +
  labs(
    title = "90th Percentile Generic Workup Duration by Complexity",
    x = "Complexity Bucket",
    y = "P90 Minutes"
  )
```
6. Generic Workup Empirical
```{r}
# Empirical workup EDA

ggplot(workup_empirical, aes(x = duration_min)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of Generic Workup Duration",
    x = "Minutes",
    y = "Count"
  )

ggplot(workup_empirical, aes(x = complexity_bucket, y = duration_min)) +
  geom_boxplot() +
  labs(
    title = "Generic Workup Duration by Complexity",
    x = "Complexity Bucket",
    y = "Minutes"
  )
```

7. Imaging Probability and Modality Mix: This file tells you whether imaging happens and, if so, what modality is used. (looks like theres nothing useful in this file?)
```{r}

# Imaging probability

# Average imaging probability by complexity
imaging_by_complexity <- imaging_prob %>%
  group_by(complexity_bucket) %>%
  summarise(avg_imaging_prob = mean(needs_imaging_prob, na.rm = TRUE), .groups = "drop")

ggplot(imaging_by_complexity, aes(x = complexity_bucket, y = avg_imaging_prob)) +
  geom_col() +
  labs(
    title = "Average Imaging Probability by Complexity",
    x = "Complexity Bucket",
    y = "Probability of Imaging"
  )

# Average imaging probability by triage
imaging_by_triage <- imaging_prob %>%
  group_by(triage_priority) %>%
  summarise(avg_imaging_prob = mean(needs_imaging_prob, na.rm = TRUE), .groups = "drop")

ggplot(imaging_by_triage, aes(x = factor(triage_priority), y = avg_imaging_prob)) +
  geom_col() +
  labs(
    title = "Average Imaging Probability by Triage Priority",
    x = "Triage Priority",
    y = "Probability of Imaging"
  )

# Reshape modality probabilities for plotting
modality_plot <- imaging_prob %>%
  select(complexity_bucket, triage_priority, xr_prob, ct_prob, mri_prob, us_prob) %>%
  pivot_longer(cols = c(xr_prob, ct_prob, mri_prob, us_prob),
               names_to = "modality",
               values_to = "probability")

ggplot(modality_plot, aes(x = modality, y = probability)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Imaging Modality Probabilities",
    x = "Modality",
    y = "Probability"
  )
```
8. Imaging Duration File: This file gives timing by modality, including acquisition, interpretation, and total imaging time.
```{r}

# Imaging duration 

# Total imaging median by modality
ggplot(imaging_duration, aes(x = imaging_type, y = total_imaging_median_min)) +
  geom_col() +
  labs(
    title = "Median Total Imaging Time by Modality",
    x = "Imaging Type",
    y = "Median Minutes"
  )

# Total imaging p90 by modality
ggplot(imaging_duration, aes(x = imaging_type, y = total_imaging_p90_min)) +
  geom_col() +
  labs(
    title = "P90 Total Imaging Time by Modality",
    x = "Imaging Type",
    y = "P90 Minutes"
  )
```



### DES Model
## Sampling Functions
```{r}
approx_weibull_from_median_p90 <- function(median_min, p90_min) {
  stopifnot(is.numeric(median_min), is.numeric(p90_min))
  stopifnot(length(median_min) == 1, length(p90_min) == 1)
  stopifnot(median_min > 0, p90_min > median_min)

  k <- log(log(10) / log(2)) / log(p90_min / median_min)
  lambda <- median_min / (log(2)^(1 / k))

  list(shape = k, scale = lambda)
}

sample_duration_weibull <- function(median_min, p90_min, n = 1) {
  pars <- approx_weibull_from_median_p90(median_min, p90_min)
  rweibull(n, shape = pars$shape, scale = pars$scale)
}
```

```{r}
## How long until next patient arrives
sample_interarrival <- function(hour, day_label) {
  row <- interarrival %>%
    filter(hour_of_day == hour,
           day_of_week_label == day_label) %>%
    slice(1)

  if (nrow(row) == 0 || is.na(row$mean_interarrival_min)) {
    return(10)  # fallback
  }

  return(rexp(1, rate = 1 / row$mean_interarrival_min))
}
```

```{r}
# Assign who the patient is, acuity, complexity, etc.
sample_attribute <- function(hour, day_num, attr) {
  probs <- case_mix %>%
    filter(hour_of_day == hour,
           day_of_week_num == day_num,
           attribute_name == attr)

  if (nrow(probs) == 0) return(NA_real_)

  val <- sample(probs$attribute_value, 1, prob = probs$probability)

  val_num <- suppressWarnings(as.numeric(val))

  if (!is.na(val_num)) return(val_num)

  as.numeric(as.factor(val))
}
```

```{r}
# First seen delay (waiting room time)
sample_first_seen <- function(triage) {
  row <- first_seen_summary %>%
    filter(triage_priority == triage) %>%
    slice(1)

  if (nrow(row) == 0) return(30)

  sample_duration_weibull(row$median_min, row$p90_min, 1)
}
```

```{r}
# Workup time (main ED time)
sample_workup <- function(complexity) {
  row <- workup_summary %>%
    filter(complexity_bucket == complexity) %>%
    slice(1)

  if (nrow(row) == 0) return(60)

  sample_duration_weibull(row$median_min, row$p90_min, 1)
}

```

```{r}
# Imaging decision (Yes/No for now since missing the imaging probability data)
sample_imaging <- function(complexity, triage) {
  row <- imaging_prob %>%
    filter(complexity_bucket == complexity,
           triage_priority == triage) %>%
    slice(1)

  if (nrow(row) == 0) return(FALSE)

  runif(1) < row$needs_imaging_prob
}
```

```{r}
# Imaging time
sample_imaging_time <- function() {
  mean(imaging_duration$total_imaging_median_min, na.rm = TRUE)
}
```

## TEST: Simulate One Patient
```{r}
simulate_patient <- function(hour, day_num, day_label) {

  # attributes
  triage <- sample_attribute(hour, day_num, "acuity")
  complexity <- sample_attribute(hour, day_num, "complexity_bucket")

  # steps
  wait_time <- sample_first_seen(triage)
  workup_time <- sample_workup(complexity)

  imaging_flag <- sample_imaging(complexity, triage)

  if (imaging_flag) {
    imaging_time <- sample_imaging_time()
  } else {
    imaging_time <- 0
  }

  total_time <- wait_time + workup_time + imaging_time

  return(data.frame(
    triage = triage,
    complexity = complexity,
    wait_time = wait_time,
    workup_time = workup_time,
    imaging = imaging_flag,
    imaging_time = imaging_time,
    total_time = total_time
  ))
}
```

```{r}
set.seed(1)
simulate_patient(14, 2, "Tuesday")
```
## TEST: Simulate Many Patients
```{r}
patients <- bind_rows(
  replicate(1000,
            simulate_patient(14, 2, "Tuesday"),
            simplify = FALSE)
)
```

```{r}
summary(patients$total_time)

ggplot(patients, aes(x = total_time)) +
  geom_histogram(bins = 40) +
  labs(title = "Simulated ED Length of Stay")
```

## Actual DES Model
```{r}
library(simmer)

env <- simmer("ED")

traj <- trajectory() %>%

  # assign attributes
  set_attribute("triage", function() sample_attribute(14, 2, "acuity")) %>%
  set_attribute("complexity", function() sample_attribute(14, 2, "complexity_bucket")) %>%

  # wait to be seen
  timeout(function() {
  triage <- get_attribute(env, "triage")
  sample_first_seen(triage)
}) %>%

  # workup
  timeout(function() {
  complexity <- get_attribute(env, "complexity")
  sample_workup(complexity)
}) %>%

  # imaging decision
  branch(
  function() {
    triage <- get_attribute(env, "triage")
    complexity <- get_attribute(env, "complexity")

    if (sample_imaging(complexity, triage)) 1 else 0
  },
  continue = c(TRUE),
  trajectory() %>%
    timeout(function() sample_imaging_time())
)
```

```{r}
env %>%
  add_generator("patient",
                traj,
                function() sample_interarrival(14, "Tuesday")) %>%
  run(until = 100)
```

